From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-09 paper/paper.Rmd
adding refs

abstract: |
  # Article Type: Review & Synthesis
  # Running title: Noise phenomena
  # Figures: 4
  # Tables: 0
  # Word count: 7488
  # Word count in abstract: 175
  # Number of References: 54
  # Archive statement:

@@

  I provide a brief synthesis of ecological research in stochastic phenomena.  Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  Yet with each aspect of stochasticity leading to some new or unexpected behavior, the time is right to move beyond the familiar refrain of "everything is important" [@Bjornstad2001].  Stochastic phenomena can suggest new ways of inferring process from pattern, and thus spark more dialog between theory and empirical perspectives that best advances the field as a whole. I highlight a few compelling examples, while observing that the study of stochastic phenomena are only beginning to make this translation into empirical inference.  There are rich opportunities at this interface in the years ahead.

@@

@@

@@

The challenge of identifying warning signs that a system is approaching
a tipping point has proven a rich area for applications of stochastic
processes.  We have already seen how in addition to the patterns of
increasing variance and auto-correlation, our survey of stochastic phenomena
has highlighted the increased potential to observe stochastic switching,
or flickering, as the system approaches the tipping point.  Challenges remain
in distinguishing these patterns from other causes [@Boettiger2013a].
For instance, as we have seen, variance in a system can increase due to
external drivers without changing the auto-correlation, and both variance
and auto-correlation can increase if a system loses stability without
approaching a saddle-node bifurcation, but will not show flickering.
The greater our suite of stochastic indicators, the better our ability
to pin down the processes involved.  Such examples suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, in which a stable node becomes
a limit cycle.

>>>>

abstract: |
  # Article Type: Review & Synthesis
  # Running title: Noise phenomena
  # Figures: 4
  # Tables: 0
  # Word count: 
  # Word count in abstract: 
  # Number of References: 
  # Archive statement:

@@

   Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  Yet with each aspect of stochasticity leading to some new or unexpected behavior, the time is right to move beyond the familiar refrain of "everything is important" [@Bjornstad2001].  Stochastic phenomena can suggest new ways of inferring process from pattern, and thus spark more dialog between theory and empirical perspectives that best advances the field as a whole. I highlight a few compelling examples, while observing that the study of stochastic phenomena are only beginning to make this translation into empirical inference.  There are rich opportunities at this interface in the years ahead.

@@

<!-- Need concise overview of stochasticity literature -->

@@

<!-- Shorten this, more citations, more focus -->

@@

The challenge of identifying warning signs that a system is approaching
a tipping point has proven a rich area for applications of stochastic
processes.  We have already seen how in addition to the patterns of
increasing variance and auto-correlation, our survey of stochastic phenomena
has highlighted the increased potential to observe stochastic switching,
or flickering, as the system approaches the tipping point.  Challenges remain
in distinguishing these patterns from other causes [@Boettiger2013a].
For instance, as we have seen, variance in a system can increase due to
external drivers without changing the auto-correlation, and both variance
and auto-correlation can increase if a system loses stability without
approaching a saddle-node bifurcation, but will not show flickering.
The greater our suite of stochastic indicators, the better our ability
to pin down the processes involved.  Such examples suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussman2000; @Strogatz2001].
